---
title: "Explore US Government R&D Spendings"
subtitle: "TidyTuesday 7 - Source AAAS"
author: null
date: '2019-02-16'
summary: "Combine ggplot2 with hierarchical and k-means clustering"
slug: government-funding
categories: []
tags: []
image:
  caption: ''
  focal_point: ''
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
    number_sections: true
    dev: svglite
    css: ../custom.css
editor_options: 
  chunk_output_type: console
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


Recently the [TidyTuesday social project proposed a dataset documenting R&D spending of the US government](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-02-12). This dataset was originally collected by the [American Association for the Advancement of Science](https://www.aaas.org/).

This tidy dataset stores R&D spending by the US government from 1976 to 2016, devided by agency. It's available [here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-02-12).

# Load Packages and Download Files

Let's explore this dataset. 

First we load the packages that we need.

```{r}
library(tidyverse)
library(scico)
```

And then the data.

```{r}
# Get data ---------------------------------------------------

rd_path <- "_data/2-07-funding.Rdata"
rd_url <- paste0("https://raw.githubusercontent.com/",
                 "rfordatascience/tidytuesday/master/data/",
                 "2019/2019-02-12/fed_r_d_spending.csv")

if(!file.exists(rd_path)) {
  rd <- 
    read_csv(rd_url)
  
  save(rd, file = rd_path)
  
} else {
  load(rd_path)
}
```

The `rd` dataset has this aspect.

```{r}
rd %>% print()
```

We are interested in the spending, which, as stated in the [data dictionary](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-02-12#total-federal-rd-spending-by-agencydeparment) are stored in the variable **rd_budget**.
Their unit is **inflation adjusted dollars**.

# Explore - Which Agencies Get Similar Funding?

Which agencies get similar fundings? Do funding to certain groups of research agencies correlate? And what could we learn from that, could we guess policies?

## Correlation between funding pattern

This was one of the first things that I explored, and I got lucky. It seems that there are two groups of correlation. 

To test correlation, we must switch to the wide data format, because the `cor()` function requires that each variable is in its own column.

```{r}
# Wide format
rd_wide <- 
  rd %>% 
  select(department, year, rd_budget) %>% 
  tidyr::spread(key = department,
                value = rd_budget)

# correlation of everything against everything
rd_wide %>%
  column_to_rownames("year") %>%
  # Test correlation
  cor() %>% 
  # visualize
  as_tibble(rownames = "agency") %>% 
  gather(DHS:VA, key = "agency_2", value = "corr") %>% 
  ggplot(aes(x = agency,
             y = agency_2,
             fill = corr)) +
  geom_tile() + 
  scale_fill_viridis_c() +
  theme(aspect.ratio = 1,
        axis.text.x = element_text(angle = 90, hjust = .5))
```

Lucky, it looks there are two defined groups of correlation. Could we capture this behaviour in clusters?

## Clusters of funding patterns?

With clusters, we want to capture the behaviour of our variables. Meaning that we want to capture how one variable behaves compared to another: when one variable increases in value, does the other do the same? 

So, to make variables comparable, we must scale them. 

Probably, the most common scaling is by z-scores. But, because z-scores are centered at 0. This introduces **negative numbers**, but spendings for R&D agencies are (hopefully ;) ) never negative.

Thus, I decided to rescale spendings for each agency **between 0 and 1**, in this way we can make them comparable and present them in a more intuitive way. Moreover, this kind of scaling **preserves 0 as anchored point**.

```{r}
# scale each variable between 0 and 1
rd_wide_01 <- 
  rd_wide %>%
  mutate_at(vars(-contains("year")),
            ~scales::rescale(., to = c(0,1),
                             from = c(0, max(.))))


# this is how it looks after scaling
rd_wide_01 %>% print()
```

After scaling, I tried **hierachical clustering**. But, clustering functions such as `hclust()` or `kmeans()`, by default, cluster the rows of a dataset: we must to modify the format of the data again. 

This time we transpose the data with `t()`, to exchange rows and columns.

```{r}
# For clustering
# we need to transpose the data
# because dist() %>% clusters the columns
rd_for_clust <-   
  rd_wide_01 %>%
  column_to_rownames("year") %>%
  t() 

rd_hclust <-
  rd_for_clust %>% 
  dist() %>%
  hclust() 
```

The results of hierarchical clustering are a bit more complex:

- The **DHS** stays by itself, since it was established late, in the '00s.
- **VA, HHS, NIH** and **HHS** group together, since lately, they have seen an increase in fundings,
- **EPA** is also pretty much by itself, since it's fundings kept decreasing.

```{r}
rd_hclust %>% plot()
```

## Visualize the clusters with heatmaps

In biology, small numbers of clustered observation are often visualized with a combination of heatmaps and dendrograms.

You can plot those together with the packages [superheat](https://rlbarter.github.io/superheat/clustering.html#dendrogram)

```{r}
rd_for_clust %>% 
  superheat::superheat(row.dendrogram = TRUE,
                       left.label.text.size = 3,
                       bottom.label.text.angle = 90,
                       bottom.label.text.size = 3,
                       grid.hline.col = "grey40",
                       grid.vline.col = "grey40")
```

# Other visualizations.

Heatmaps are great for the tight space that scientific journals reserve to figures. 

But, here, because I don't have limitations and for fun, I wanted to try other visualization.

The one below has more white space, but might be more pleasurable to the eye than an highly compressed heatmap, and it conveys additional information, such as maximum absolute spending for each department

```{r}
# I restarted from the original data,
# to have both scaled and absolute measurements
#  in the same dataset
rd_to_plot <- 
  rd %>% 
  select(department, year, rd_budget) %>% 
  # rescale 0 to 1
  group_by(department) %>% 
  mutate(budget_01 = rd_budget %>% 
           {scales::rescale(., to = c(0, 1),
                            from = c(0, max(.)))}) %>% 
  ungroup() 
```

I would like to prepare a faceted plot, in which facets are ordered based on a dendrogram, and with the dendrogram itself on the side.

I can order the facets with the dendrogram from hierarchical clustering using a factor. (But I have no idea how to show the dendrogram itself.)

```{r} 
rd_to_plot <- 
  rd_to_plot %>% 
  mutate(department = department %>% 
           factor(levels = rd_hclust$labels[rd_hclust$order]))
```

```{r, fig.height=10}
line_col <- "grey70"
text_col <- "grey40"
line_size <- .2
label_spacer <- .4

p <- 
  rd_to_plot %>% 
  ggplot(aes(x = year,
             y = budget_01,
             colour = budget_01)) +
  geom_line(colour = line_col,
            size = line_size) +
  geom_segment(data = . %>% 
              group_by(department) %>% 
              filter(rd_budget == max(rd_budget)),
            aes(y = budget_01,
                yend = budget_01 -label_spacer,
                xend = year),
            size = line_size, colour = line_col) +
  geom_point(aes(size = rd_budget)) +
  geom_text(data = . %>% 
              group_by(department) %>% 
              filter(rd_budget == max(rd_budget)),
            aes(y = budget_01 - label_spacer - .1,
                label = rd_budget %>% scales::scientific(digits = 0)),
            size = 4) +
  facet_grid(department ~ .) +
  theme_minimal() + 
  scale_color_scico(palette = "devon",
    guide = FALSE, 
    direction = -1,
    begin = .2) +
  lims(y = c(0, 1.1)) +
  labs(y = "Government R&D Spending") +
  scale_size(range = c(.1, 3), guide = FALSE) +
  theme(text = element_text(family = "sans",
                            colour = text_col),
        axis.text.y = element_blank(),
        panel.grid = element_blank(),
        panel.grid.major.x = element_line(colour = line_col,
                                          size = line_size/2),
        legend.position = "top",
        strip.text.y = element_text(family = "sans",
                                    angle = 0,
                                    colour = text_col))


p %>% print()
```

What do you think?


